Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking

### 2023-07-14 manuscript/manuscript.qmd
first draft of manuscript done.

While laboratory-based polysomnography (PSG) is typically considered the
gold standard for objective sleep measurement, its practicality in
large-scale epidemiological studies is hindered due to high costs, the
necessity for professional administration, and it is also subject to
potential rater bias[@vandewater2011; @lee2022]. As an alternative,
diaries are commonly employed as cost-effective and low-tech methods for
sleep assessment in population research. However, reliance on
diary-based methods may lead to recall bias and other
limitations[@moore2015]. A more feasible approach in large-scale
epidemiological studies is to use device-based measurement methods that
can estimate sleep duration. This approach offers the advantage of being
less burdensome for participants and eliminates potential biases
associated with recall.

Body-worn accelerometers have emerged as an effective and affordable
alternative for objectively assessing sleep patterns at home over
extended periods. These devices gather continuous, high-resolution data
for several weeks without the need for recharging, thus reducing
participant burden. Initial applications of accelerometry for sleep and
wake stage classification were based on wrist movements, starting with
an algorithm developed in 1982 and validated with PSG [@webster1982].
This model was later refined in 1992 [@cole1992], giving rise to the
widely used Cole-Kripke model. As the field advanced, an array of
techniques, including heuristic algorithms, machine learning models,
regression, and deep learning, were employed to analyze hip
andwrist-worn accelerometer data[@palotti2019; @cole1992; @sazonov2004;
@sadeh1994; @hees2015; @sundararajan2021].

@@

## Algorithms, Training and Validation

In this study, we employed two distinct learning strategies to analyze
sleep patterns from thigh-mounted accelerometer data. We used a
sequential strategy, comprising an ensemble of four pairs of models,
each pair featuring the same algorithm. This strategy aimed to make the
prediction task more manageable for the algorithms by breaking it down
into a sequence of two binary classifications: first predicting 'in-bed'
time, then 'sleep' time. Simultaneously, we also used a multiclass
approach utilizing a bidirectional Long Short-Term Memory
(biLSTM)[@hochreiter1997] neural network.

### Models in Sequence

To predict in-bed time and sleep time accurately, we employed an
ensemble learning strategy based on sequential binary classification
models. This approach involved constructing a sequence of models using
multiple machine learning algorithms to improve predictive accuracy. The
process began with an initial model predicting in-bed time, followed by
a second model that utilized the output of the initial model to predict
sleep time. This sequential approach was applied across all four
algorithms detailed below, with each subsequent model leveraging the
outputs of the previous models for improved predictions.

@@

### Multiclass Model

We also employed a biLSTM, a multiclass classifier, to predict three
distinct states: out-of-bed-awake, in-bed-awake, and in-bed-asleep. The
architecture of the biLSTM was set up with four layers, each equipped
with 128 hidden units. This configuration was intentionally chosen to
balance between model complexity and training efficiency: it provided
the depth necessary for learning intricate patterns while remaining
feasible for timely training. The bidirectional design of the LSTM
served to enhance data interpretation and mitigate overfitting by
doubling the hidden units at each time step. For input, we used tensors
shaped as sequences, with each sequence spanning 10 minutes and a step
size of one.This approach follows in the footsteps of previous studies
that utilized LSTM models for sleep detection. These studies showcased
the promising potential of LSTMs in capturing complex temporal patterns.
Particularly, the works of Sano et al. [@sano2019] and Chen et al.
[@chen2021] demonstrated the effectiveness of LSTM models in improving
sleep detection using accelerometer data, underscoring the value of this
modeling approach.

### Model Training

For the models in sequence, we trained four pairs of classification
models. Each pair was designed to distinguish between in-bed/out-of-bed
and asleep/awake states, respectively. The dataset was randomly split
into a training set and a testing set, each containing approximately 50%
of the subjects. This division was ensured to not have samples from the
same night simultaneously present in both sets. To optimize
hyperparameters, we performed a 10-fold Monte Carlo cross-validation on
a regular grid, comprising 20 different combinations of hyperparameters.
The F1 score served as the optimization metric. The best-performing set
of hyperparameters was then used to fit the models to the full training
dataset. This approach allowed us to maximize performance by leveraging
all available data. Moreover, after extracting the in-bed time from the
initial sequential models, the imbalance on the resulting dataset could
cause biases during model training, as models may favor predicting the
majority class. To rectify this, we employed the Synthetic Minority
Over-sampling Technique (SMOTE)[@chawla2002]. SMOTE generates new
samples by interpolating random samples with their nearest neighbors. We
utilized the themis R package[@themis] to implement SMOTE, resulting in
a balanced distribution of training samples across both classes.

@@

### Model Validation

@@

1.  Sleep Period Time (SPT) - This refers to the total duration of the
    sleep period, which is defined as the time from the start to the end
    of the ZM recording.

@@

A complete set of confusion matrices generated from data both containing
the out-of-bed and in-bed time are presented in @fig-conf_mat. The
figure shows favorable epoch-to-epoch performance across across all
sequential models, however, it is evident that the biLSTM is less
successful in classifying the in-bed-awake class which cannot be deduced
from the confusion matrices from the sequential models.

@@

@tbl-ba_cor presents a comparative analysis of the included models used
to predict various sleep quality parameters (SPT, TST, SE, LPS, WASO)
using the 5-minute median filtered ZM predictions. To see the full table
including models developed from raw ZM predictions and 10-minute median
filtered ZM predictions, see **\[SUPP. MAT.\]**. In terms of bias, the
decision tree model consistently underestimated SPT, TST, and SE, and
overestimated LPS and WASO in comparison to ZM. The logistic regression
model had similar trends, with more pronounced underestimation in TST
and overestimation in LPS. The eed-forward neural network also exhibited
similar bias as the decision tree and the logistic regression models,
but with a higher overestimation in WASO. On the other hand, the XGBoost
model showed least bias among all, especially in its 5-minute median
predictions. Considering LOA, the decision tree had higher variability
across different sleep quality parameters and filtering techniques,
particularly for LPS and WASO, which indicates lower agreement with ZM.
Other models had comparable LOA but with notable exceptions. For
example, TST LOA for the logistic regression model was particularly wide
in the 5-minute median predictions. Correlation-wise, the pearson
coefficient, revealed that the XGBoost model consistently had the
highest correlation with ZM across all sleep qualityparameters and
filtering methods Notably, the XGBoost's 5-minute median predictions
showed the strongest correlation (0.66) for TST among all models and
filtering techniques.

@@

@fig-xgb_ba_cor shows the agreement between the XGBoost model, trained
on 5-minute median filtered ZM predictions, and the 5-minute
median-smoothed ZM-derived sleep quality parameters. The Bland-Altman
plot for the SPT reveals a significant level of agreement with the ZM,
as evidenced by a positive correlation. However, the presence of extreme
outliers widens the limits of agreement (LOA). The scatterplot for SPT
also demonstrates a positive trend, indicating a moderate linear
correlation between the XGBoost model and the ZM-derived sleep quality
parameters.In terms of TST, perfect agreement is not observed for a
substantial number of nights, but there is a positive correlation
between the XGBoost model and ZM-derived sleep quality parameters. The
bias and LOA for TST are comparable to those observed for SPT,
indicating a consistent level of agreement between the two methods. The
scatterplot for TST also shows a slightly higher correlation, primarily
driven by the absence of extreme outliers.Furthermore, the remaining
three sleep quality parameters, SE, LPS, and WASO, exhibit
heteroscedasticity in contrast to SPT and TST. This outcome is expected
as achieving 100% sleep efficiency is relatively rare, resulting in less
disagreement between the methods as values approach the upper limit.
However, as sleep efficiency decreases, the potential for discrepancies
and differing interpretations between the methods increases, leading to
greater heteroscedasticity. A moderate linear correlation is observed
between the XGBoost model and ZM-derived sleep quality parameters for
SE, indicating a positive relationship. However, a poor correlation is
observed for LPS and WASO, suggesting less agreement between the two
methods for these parameters. Similar plots for all models are available
in the supplementary materials.

@@

In our evaluation of sleep quality parameters, we found that Latency to
Persistent Sleep (LPS) had the largest mean error relative to absolute
time allocated to LPS. This suggests that the initial epochs of Sleep
Period Time (SPT) are particularly challenging to classify correctly.
This is also supported by the poor Pearson correlations between LPS
derived from model predictions and the ZM. The XGBoost model, which was
the best performer among all models, overestimated LPS by an average of
26.4 minutes for models trained on raw ZM predictions, 28.5 minutes for
models trained on 5-minute filtered ZM predictions, and 34.5 minutes for
models trained on 10-minute filtered ZM predictions. This level of
discrepancy is comparable to the mean error of sleep latency of 23
minutes reported by Johansson et al.[@johansson_development_2023].
Johansson et al. suggest that the discrepancy with the gold standard is
likely due to the multifaceted nature of the sleep state, which is a
complex physiological process. Short awakenings or sleep episodes may
not necessarily correspond to noticeable changes in thigh movement,
making them difficult to detect and accurately classify. These results
align with several methods for wrist-worn devices reviewed by Conley and
colleagues[@conley2019]. They reported correlations between
accelerometer and PSG sleep onset latency (equivalent to LPS) from 10
studies with a mean correlation of 0.2 (ranging from -0.69 to 0.69),
indicating the inherent difficulty in estimating this parameter using
accelerometry alone.

@@

The study boasts several strengths, including the capacity to derive
in-bed awake and asleep times from out-of-bed data, thereby allowing for
the extraction of vital sleep quality parameters. Furthermore, the
research benefits from evaluating multiple nights per subject, providing
valuable information into intra-subject sleep variability. However,
certain limitations exist. The use of ZM, which isn't recognized as a
gold standard, could potentially compromise our findings' validity.
Future research could consider using PSG as a reference for methods
similar to ours, despite its limitations, for a more accurate
comparison. Moreover, our models weren't validated using an external
dataset, a process that would have showcased their broader
applicability. Hence, our conclusions remain confined primarily to
children.

In conclusion, our study contributes to the ongoing efforts to improve
sleep estimation methods using thigh-worn accelerometers. We evaluated
different machine learning models and configurations for predicting
in-bed and sleep times and their corresponding sleep quality parameters.
While the sequential models generally demonstrated excellent performance
in predicting in-bed time, they faced challenges in accurately
distinguishing between sleep and wake epochs during in-bed time.
Nevertheless, their performance could be substantially improved through
the application of median filters. Among all models and configurations
evaluated, the XGBoost model exhibited superior performance across all
performance metrics, including epoch-to-epoch predictions and sleep
quality parameter derivatives. Our research also highlighted the current
limitations of sleep detection methods, such as challenges in
effectively detecting wake periods during in-bed time and the need for
further improvements to increase the precision of individual sleep
assessments. We believe our work lays the groundwork for future research
to further refine and improve the performance of these models,
contributing to a more precise and accurate evaluation of sleep patterns
and quality using thigh-worn accelerometers.

>>>>

The traditional gold standard for objective sleep measurement,
laboratory-based polysomnography (PSG), has been found to be impractical
in large-scale epidemiological studies due to its high cost, need for
professional administration, and susceptibility to rater
bias[@vandewater2011; @lee2022]. As an alternative, diaries have been
used due to their cost-effectiveness and simplicity, although they are
subject to recall bias and other limitations[@moore2015]. An innovative
approach involves device-based measurement methods. These tools, which
estimate sleep duration, are advantageous due to their reduced
participant burden and elimination of potential recall biases. A
prominent example of such tools is body-worn accelerometers, which offer
a practical and affordable means of objectively assessing sleep patterns
at home for extended periods. Accelerometers collect continuous,
high-resolution data for several weeks without requiring recharging,
further minimizing participant burden. Their use in sleep and wake
classification began with a wrist movement-based algorithm developed in
1982, and validated using PSG[@webster1982]. This algorithm was refined
in 1992[@cole1992], leading to the widely adopted Cole-Kripke model.
With advancements in the field, a variety of techniques, including
heuristic algorithms, machine learning models, regression, and deep
learning, are now used to analyze data from hip and wrist-worn
accelerometers[@palotti2019; @cole1992; @sazonov2004; @sadeh1994;
@hees2015; @sundararajan2021].

@@

## Algorithms

In this study, we employed two distinct learning strategies to analyze
sleep patterns from thigh-mounted accelerometer data. The first strategy
employed a sequential methodology, using an ensemble of four pairs of
models, with each pair featuring the same algorithm. This sequential
strategy was designed to simplify the prediction task by decomposing it
into two stages: first predicting 'in-bed' time, then 'sleep' time. The
output from the first set of models, which predicted in-bed time, was
subjected to a 5-minute median filter to eliminate leading and trailing
in-bed time blips. This process enabled us to extract a single
continuous time interval that we identified as the sleep period time
window (SPT). The SPT, in turn, served as the input to the second models
in the sequence, enhancing their predictive accuracy for sleep time. We
applied this sequential strategy across all four algorithms, as detailed
below.

@@

In parallel, we also employed a multiclass approach using a
bidirectional Long Short-Term Memory (biLSTM)[@hochreiter1997] neural
network. This network, which was designed to predict three distinct
classes: 'out-of-bed-awake', 'in-bed-awake', and 'in-bed-asleep', was
configured with four layers and 128 hidden units per layer. This balance
between model complexity and training efficiency was intended to
facilitate learning of intricate patterns while ensuring feasible
training times. The bidirectional nature of the LSTM enhanced data
interpretation and reduced overfitting by doubling the hidden units at
each time step. The LSTM model used sequences of tensors as input, with
each sequence spanning 10 minutes and a step size of one. As
demonstrated by previous studies such as those by Sano et al.
[@sano2019] and Chen et al. [@chen2021], LSTM models have shown great
promise in sleep detection using accelerometer data, thanks to their
ability to capture complex temporal patterns.

## Model Training

For the models in sequence, we trained four pairs of classification
models. Each pair was designed to distinguish between in-bed/out-of-bed
and asleep/awake states, respectively. The dataset was randomly split
into a training set and a testing set, each containing approximately 50%
of the subjects. The splitting of the data was ensured to not have
samples from the same night simultaneously present in both sets. To
optimize hyperparameters, we performed a 10-fold Monte Carlo
cross-validation on a regular grid, comprising 20 different combinations
of hyperparameters. The F1 score served as the optimization metric. The
best-performing set of hyperparameters was then used to fit the models
to the full training dataset. This approach allowed us to maximize
performance by leveraging all available data. Moreover, after extracting
the in-bed time from the initial sequential models, the imbalance on the
resulting dataset could cause biases during model training, as models
may favor predicting the majority class. To rectify this, we employed
the Synthetic Minority Over-sampling Technique (SMOTE)[@chawla2002].
SMOTE generates new samples by interpolating random samples with their
nearest neighbors. We utilized the themis R package[@themis] to
implement SMOTE, resulting in a balanced distribution of training
samples across both classes.

@@

## Model Validation

@@

1.  Sleep Period Time (SPT) - This refers to the total duration of time
    in bed with the intention to sleep, which is defined as the time
    from the start to the end of the ZM recording.

@@

A complete set of confusion matrices generated from data both containing
the out-of-bed and in-bed time are presented in @fig-conf_mat. The
figure shows favorable epoch-to-epoch performance across all sequential
models, however, it is evident that the biLSTM is less successful in
classifying the in-bed-awake class which cannot be deduced from the
confusion matrices from the sequential models.

@@

@tbl-ba_cor presents a comparative analysis of the included models used
to predict various sleep quality parameters (SPT, TST, SE, LPS, WASO)
using the 5-minute median filtered ZM predictions. To see the full table
including models developed from raw ZM predictions and 10-minute median
filtered ZM predictions, see **\[SUPP. MAT.\]**. In terms of bias, the
decision tree model consistently underestimated SPT, TST, and SE, and
overestimated LPS and WASO in comparison to ZM. The logistic regression
model had similar trends, with more pronounced underestimation in TST
and overestimation in LPS. The feed-forward neural network also
exhibited similar bias as the decision tree and the logistic regression
models, but with a higher overestimation in WASO. On the other hand, the
XGBoost model showed least bias among all, especially in its 5-minute
median predictions. Considering LOA, the decision tree had higher
variability across different sleep quality parameters and filtering
techniques, particularly for LPS and WASO, which indicates lower
agreement with ZM. Other models had comparable LOA but with notable
exceptions. For example, TST LOA for the logistic regression model was
particularly wide in the 5-minute median predictions. Correlation-wise,
the pearson coefficient, revealed that the XGBoost model consistently
had the highest correlation with ZM across all sleep quality parameters
and filtering methods Notably, the XGBoost's 5-minute median predictions
showed the strongest correlation (0.66) for TST among all models and
filtering techniques.

@@

@fig-xgb_ba_cor shows the agreement between the XGBoost model, trained
on 5-minute median filtered ZM predictions, and the 5-minute
median-smoothed ZM-derived sleep quality parameters. The Bland-Altman
plot for the SPT and TST reveals a significant level of agreement with
the ZM, as evidenced by a bias close to zero. Interestingly, a portion
of the data points are located near the zero line indicating perfect
agreement. However, the presence of extreme outliers widens the limits
of agreement (LOA). The scatterplot for SPT also demonstrates a positive
trend, indicating a moderate linear correlation between the XGBoost
model and the ZM-derived sleep quality parameters. The bias and LOA for
TST are comparable to those observed for SPT, indicating a consistent
level of agreement between the two methods. The scatterplot for TST also
shows a slightly higher correlation, primarily driven by the absence of
extreme outliers.Furthermore, the remaining three sleep quality
parameters, SE, LPS, and WASO, exhibit heteroscedasticity in contrast to
SPT and TST. This outcome is expected as achieving 100% sleep efficiency
is relatively rare, resulting in less disagreement between the methods
as values approach the upper limit. However, as sleep efficiency
decreases, the potential for discrepancies and differing interpretations
between the methods increases, leading to greater heteroscedasticity. A
moderate positive linear correlation is observed between the XGBoost
model and ZM-derived sleep quality parameters for SE, however, a poor
correlation is observed for LPS and WASO, suggesting less agreement
between the two methods for these parameters. Similar plots for all
models are available in the **supplementary materials (75 plots. Jeg ved
ikke om nogen orker at bladre igennem dem).**

@@

In our evaluation of sleep quality parameters, we found that LPS had the
largest mean error relative to absolute time allocated to LPS. This
suggests that the initial epochs of Sleep Period Time (SPT) are
particularly challenging to classify correctly. This is also supported
by the poor Pearson correlations between LPS derived from model
predictions and the ZM. The XGBoost model, which was the best performer
among all models, overestimated LPS by an average of 26.4 minutes for
models trained on raw ZM predictions, 28.5 minutes for models trained on
5-minute filtered ZM predictions, and 34.5 minutes for models trained on
10-minute filtered ZM predictions. This level of discrepancy is
comparable to the mean error of sleep latency of 23 minutes reported by
Johansson et al.[@johansson_development_2023]. Johansson et al. suggest
that the discrepancy with the gold standard is likely due to the
multifaceted nature of the sleep state, which is a complex physiological
process. Short awakenings or sleep episodes may not necessarily
correspond to noticeable changes in thigh movement, making them
difficult to detect and accurately classify. These results align with
several methods for wrist-worn devices reviewed by Conley and
colleagues[@conley2019]. They reported correlations between
accelerometer and PSG sleep onset latency (equivalent to LPS) from 10
studies with a mean correlation of 0.2 (ranging from -0.69 to 0.69),
indicating the inherent difficulty in estimating this parameter using
accelerometry alone.

@@

The study boasts several strengths, including the capacity to
distinguish in-bed awake and asleep from out-of-bed, thereby allowing
for the extraction of vital sleep quality parameters. Furthermore, the
research benefits from evaluating multiple nights per subject, providing
valuable information into intra-subject sleep variability. However,
certain limitations exist. The use of ZM, which isn't recognized as a
gold standard, could potentially compromise our findings' validity.
Future research could consider using PSG as a reference for methods
similar to ours, despite its limitations, for a more accurate
comparison. Moreover, our models weren't validated using an external
dataset, a process that would have showcased their broader
applicability. Hence, our conclusions remain confined primarily to
children.

In conclusion, our study contributes to the ongoing efforts to improve
sleep estimation methods using thigh-worn accelerometers. We evaluated
different machine learning models for predicting in-bed and sleep times
and their corresponding sleep quality parameters. While the sequential
models generally demonstrated excellent performance in predicting in-bed
time, they faced challenges in accurately distinguishing between sleep
and wake epochs during in-bed time. Among all models and configurations
evaluated, the XGBoost model exhibited superior performance across all
performance metrics, including epoch-to-epoch predictions and sleep
quality parameter derivatives. Our research also highlighted the current
limitations of sleep detection methods, such as challenges in
effectively detecting wake periods during in-bed time and the need for
further improvements to increase the precision of individual sleep
assessments. We believe our work lays the groundwork for future research
to further refine and improve the performance of these models,
contributing to a more precise and accurate evaluation of sleep patterns
and quality using thigh-worn accelerometers.
